Parameterization of the responses of subarctic European vegetation to key environmental variables for ozone risk assessment

### 2022-05-07 VegOzone.tex
Rephrase and mark passages that need rephrasing after new results

\begin{table}[t]
  \caption{Start and end of growing season in \unit{doy}.}
  \label{tab:sensitivity_tests_gs}
  \begin{tabular}{lcll}
    \tophline
    Species & Year & $A_\mathrm{start}$ & $A_\mathrm{end}$\\
    \middlehline
    Central Europe$^b$ & -- & 100 & 307\\
    Latitude model ($70\,\unit{^\circ N}$)$^b$ & -- & 135 & 257\\
    \multirow{2}{*}{Deciduous tree} & 2018 & 129$^*$ & 261$^a$ \\
    & 2019 & 130$^c$ & 274$^a$ \\
    \multirow{2}{*}{Coniferous tree} & 2018 & 122$^a$ & 261$^a$ \\
    & 2019 & 106$^a$ & 274$^a$ \\
    \multirow{2}{*}{Perennial grassland} & 2018 & 159$^d$ & 261$^a$\\
    & 2019 & 161$^d$ & 274$^a$ \\
    \bottomhline
  \end{tabular}
  \belowtable{$^a$ MODIS (Aqua/Terra) GPP product;\\
    $^b$ \citet{ICP:MappingManual2017},\\
    $^c$ $5\,\unit{days}-5\,\unit{^\circ C}$-rule, $T_\mathrm{air}$ from \href{seNorge.no}{seNorge.no};\\
    $^d$ One month after snowmelt; reference station Øvre Neiden (Sør-Varanger, NOR)} % Table Footnotes
\end{table}

@@

The results are comprehensively displayed in Fig.~\ref{fig:pody_rel}. Critical levels for reduction in biomass \citep[deciduous forest $4\,\unit{\%}$, coniferous forest $2\,\unit{\%}$, and grasslands $10\,\unit{\%}$ biomass reduction adopted from][]{ICP:MappingManual2017,ESPR:Hayes2021} are shown as horizontal dashed lines. The different background colors indicate the temperature response function localization. The light response function localizations are denoted by $+$ (PPFD1.2) and $-$ (PPFD0.8). Circles indicate simulations with localized GS, canopy height, and leaf dimensions, whereas squares indicate simulations with MM GS for Central Europe. We chose this to demonstrate the impact of a projected prolonged GS in the future. With open symbols, the effect on $\mathrm{POD_y}$ if SWP is taken into account is displayed.

We identified that an increasing cold tolerance (MM $\rightarrow$ cold $\rightarrow$ subarctic) leads to higher ozone uptake in both probed years, while using locally adapted GS causes a reduction. Therefore, an earlier and longer GS in 2019 leads to an increased ozone uptake compared to 2018 despite more episodes of elevated \chem{[O_3]} in 2018 (see Section~\ref{subsec:anomalies} for details).

An earlier opening of stomata at low light intensities (PPFD0.8) will cause an increase in $\mathrm{POD_1}$ while a later opening (PPFD1.2) will reduce $\mathrm{POD_1}$. The conditions in 2018 reduce $\mathrm{POD_1}$ more strongly for the PPDF0.8 light response function localization indicating an increase in sensitivity to drought stress (compare PPFD0.8 and PPFD1.2 in Fig.~\ref{fig:pody_rel}a, subarctic, 2018, with SWP effect on). Due to the shape of $f_\mathrm{light}$, a symmetric variation leads to an asymmetric response in $\mathrm{POD_1}$.

For deciduous trees (Fig.~\ref{fig:pody_rel}a) all $\mathrm{POD_1}$ simulations exceed the CL by $5-25\,\unit{mmol\,\chem{O_3}\,m^{-2}}$. Among all PFTs, deciduous trees display the largest sensitivity in $\mathrm{POD_1}$. Compared to MM $\Delta\mathrm{POD_1}=(10-31)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$. In particular, the sensitivity to the $f_\mathrm{light}$ is pronounced (extent of $\angle(+,0,-)$). With MM GS, $\mathrm{POD_1}$ values were higher in 2018 than in 2019. However, taking the localized GS into account, this is reversed for the subarctic temperature response function. This can be explained by more favorable growing conditions in 2019. Low soil water potential reduced the $\mathrm{POD_1}$ values in 2018 in some cases when using the MM GS, but did not or only slightly when using the localized GS. This indicates (frost) drought conditions early in the regular GS in 2018 when the ground was still covered in snow in reality. The difference in $\mathrm{POD_1}$ due to the localization of $f_\mathrm{T}$ for the localized GS ranges between $(1.4-3.6)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(2.7-7.7)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. The maximum difference in $\mathrm{POD_1}$ due to the $f_\mathrm{light}$ localization for the PPFD0.8 (superscript) and PPFD1.2 (subscript) amount to $^{+1.6}_{-2.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ for 2018 and $^{+1.9}_{-2.7}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ for 2019.

For coniferous trees (Fig.~\ref{fig:pody_rel}b), most simulations slightly exceed the CL ($0-10\,\unit{mmol\,\chem{O_3}\,m^{-2}}$). The MM Boreal Norway spruce parameterization was already well acclimated to the prevailing temperatures and therefore differs the least from our localized parameterizations. Taking the localized GS and canopy height into account, we find a higher ozone uptake in 2019 than in 2018. The maximum difference in $\mathrm{POD_1}$ determined between localized GS for subarctic and cold parameterizations with respect to MM ranges between $(0.5-1.0)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(2.0-3.3)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. The maximum difference determined from PPFD0.8 (superscript) and PPFD1.2 (subscript) $f_\mathrm{light}$ amounts to $^{+1.0}_{-0.8}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ for 2018 and $^{+1.3}_{-1.0}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ for 2019.

For perennial grassland (Fig.~\ref{fig:pody_rel}c), all simulations with localize GS stay below the CL. Perennial grassland shows the smallest ozone dose, but a similarly large response to the temperature parameterization localizations as deciduous trees. Again, we find a reversal for predicted ozone risk in 2018 and 2019 for the subarctic type. Perennial grassland shows the lowest sensitivity to the light threshold and SWP is only relevant without localized GS (see above). The overall difference in $\mathrm{POD_1}$ determined from the localized GS for subarctic and cold types with respect to MM ranges between $(2.2-3.4)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(4.4-6.4)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. The lower maximum temperature tolerance probably caused the lower $\mathrm{POD_1}$ values calculated for 2018 using the subarctic compared to cold localization, whereas this was not seen for 2019. The maximum difference in $\mathrm{POD_1}$ determined from the PPFD0.8 (superscript) and PPFD1.2 (subscript) variation of $f_\mathrm{light}$ amount to $^{+0.3}_{-0.2}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ and $^{+0.3}_{-0.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$, respectively.

The maximum difference in $\mathrm{POD_1}$ between simulations with the MM parameterizations and the localized temperature and light response functions is of the same order of magnitude as the interannual difference and should not be neglected.
The magnitude of all described effects on $\mathrm{POD_1}$ varies between PFTs as well as years, but the predicted ozone dose for the localized $f_\mathrm{T}$ is always larger than for the MM parameterizations. This indicates a probable underestimation of ozone risk for subarctic species when using the MM parameterizations.

@@

We estimated the biomass reductions in accordance to the CLs \citep{ICP:MappingManual2017, ESPR:Hayes2021} for the most extreme temperature parameterization localization (\emph{subarctic}) for each PFT. The estimates are listed in Table~\ref{tab:biomass_reduction}. The reported range corresponds to the maximum difference in $\mathrm{POD_1}$ defined by the $\pm 20\,\unit{\%}$ variation of $f_\mathrm{light}$ with SWP=on. The total biomass of deciduous trees and perennial grassland was notablly reduced in comparison with the MM parameterization in both years. The CL for perennial grassland is higher than for the other species. From an agro-economical point of view in our focus area, where the grass is cut only once a year, a loss of biomass of more than $10\,\unit{\%}$ can have severer consequences compared to more productive areas in central Europe and may need adjustment.

Coniferous trees are the least affected and the magnitude of total biomass reduction is almost independent of the choice of the parameterization within given uncertainties. This may reflect that the temperature parameterization in the MM is rarely a limiting factor for stomatal conductance.

\begin{table}[t]
  \caption{Estimated total biomass reduction in \unit{\%} for temperature acclimation \emph{subarctic} with bespoke GS and SWP=off. The uncertainty ranges reported here correspond to maximum a divergence deducted from varying $f_\mathrm{light}$ and SWP=on. For comparison, the corresponding biomass reduction for the generic default MM averaged over both years is shown. The reported standard deviation is computed from all sensitivity simulations ($\pm 20\,\unit{\%}$ in extent of stomatal opening at low light, SWP on/off, and bespoke GS).}
  \label{tab:biomass_reduction}
\begin{tabular}{lllll}
\tophline
Year & \multicolumn{4}{c}{PFT}\\
& Deciduous tree& Coniferous tree & \multicolumn{2}{c}{Perennial grassland}\\
& & & total & above ground\\
\middlehline
2018 & $15.5^{+(1.9...2.1)}_{-(0.8...1.5)}$ & $2.5^{+0.2}_{-0.2}$ & $9.7^{+0.2}_{-0.2}$ & $12.4^{+0.2}_{-0.2}$\\
2019 & $17.4^{+2.5}_{-1.8}$ & $3.0^{+0.2}_{-0.3}$ & $10.5^{+(0.1...0.2)}_{-(0.0...0.1)}$ & $13.7^{+(0.1...0.3)}_{-(0.1...0.3)}$\\
\middlehline
$\left<\mathrm{MM}\right>$ & $11.2\pm 1.1$ & $2.31\pm 0.04$ & $7.5\pm 0.9$ & $9.6\pm 1.4$\\
\bottomhline
\end{tabular}
%\belowtable{$^*$ above ground biomass} % Table Footnotes
\end{table}

>>>>

\begin{table}[t]
  \caption{Start and end of growing season in \unit{doy}. Central Europe is shown for comparison.}
  \label{tab:sensitivity_tests_gs}
  \begin{tabular}{lcll}
    \tophline
    Species & Year & $A_\mathrm{start}$ & $A_\mathrm{end}$\\
    \middlehline
    Central Europe$^b$ & -- & 100 & 307\\
    Latitude model ($70\,\unit{^\circ N}$)$^b$ & -- & 135 & 257\\
    \multirow{3}{*}{Deciduous tree} & 2018 & 129$^*$ & 261$^a$ \\
    & 2019 & 130$^c$ & 274$^a$ \\
    & 2100 & 116$^e$ & 286$^e$ \\
    \multirow{2}{*}{Coniferous tree} & 2018 & 122$^a$ & 261$^a$ \\
    & 2019 & 106$^a$ & 274$^a$ \\
    \multirow{2}{*}{Perennial grassland} & 2018 & 159$^d$ & 261$^a$\\
    & 2019 & 161$^d$ & 274$^a$ \\
    \bottomhline
  \end{tabular}
  \belowtable{$^a$ MODIS (Aqua/Terra) GPP product;\\
    $^b$ \citet{ICP:MappingManual2017},\\
    $^c$ $5\,\unit{days}-5\,\unit{^\circ C}$-rule, $T_\mathrm{air}$ from \href{seNorge.no}{seNorge.no};\\
    $^d$ One month after snowmelt; reference station Øvre Neiden (Sør-Varanger, NOR)\\
    $^e$ Future projection based on linear regression of $^c$ (Fig.~\ref{appendix:growing_season})
    } % Table Footnotes
\end{table}

@@

The results are comprehensively displayed in Fig.~\ref{fig:pody_rel}. Critical levels for reduction in biomass \citep[deciduous forest $4\,\unit{\%}$, coniferous forest $2\,\unit{\%}$, and grasslands $10\,\unit{\%}$ biomass reduction adopted from][]{ICP:MappingManual2017,ESPR:Hayes2021} are shown as horizontal dashed lines. The different background colors indicate the temperature response function localization. The light response function localizations are denoted by $+$ (PPFD1.2) and $-$ (PPFD0.8). Circles indicate simulations with localized GS, canopy height, and leaf dimensions, TODO: (whereas squares indicate simulations with MM GS for Central Europe. We chose this to demonstrate the impact of a projected prolonged GS in the future.) With open symbols, the effect on $\mathrm{POD_y}$ with SWP taken into account is displayed.

We identified in general that an increasing cold tolerance represented by a shift in the temperature response function (MM $<$ cold $<$ subarctic) leads to an increase in $\mathrm{POD_1}$ in both probed years. TODO: (while using locally adapted GS causes a reduction.) An earlier start and overall longer GS in 2019 leads to higher $\mathrm{POD_1}$ compared to 2018 despite the occurrence of more episodes of elevated \chem{O_3} in 2018 than in 2019 (see Section~\ref{subsec:anomalies} for details). Taking into account SWP, $\mathrm{POD_1}$ was effectively only reduced for deciduous trees in 2018, while in all other cases simulation results with and without SWP effects were identical. An earlier opening of stomata at low light intensities (PPFD0.8) caused an increase in $\mathrm{POD_1}$ compared to MM while a later opening (PPFD1.2) reduced $\mathrm{POD_1}$. Due to the shape of $f_\mathrm{light}$, a symmetric variation can lead to an asymmetric response in $\mathrm{POD_1}$.

For deciduous trees (Fig.~\ref{fig:pody_rel}a) all $\mathrm{POD_1}$ simulations exceed the CL by TODO: ($5-25\,\unit{mmol\,\chem{O_3}\,m^{-2}}$) and display the largest spread in $\mathrm{POD_1}$ with respect to $f_\mathrm{light}$. In 2019, the lowest estimated $\mathrm{POD_1}$ (MM, PPFD1.2) amounts to $10\,\unit{mmol\,\chem{O_3}\,m^{-2}}$, while the highest estimate (subarctic, PPFD0.8) reaches about $20\,\unit{mmol\,\chem{O_3}\,m^{-2}}$. In 2018, this uncertainty range is much smaller ($(12-18)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$). In particular, the impact of $f_\mathrm{light}$ is pronounced (extent of $\angle(+,0,-)$). The maximum difference in $\mathrm{POD_1}$ due to the $f_\mathrm{light}$ localization for the PPFD0.8 (superscript) and PPFD1.2 (subscript) amount to $^{+1.6}_{-2.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ for 2018 and $^{+1.9}_{-2.7}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ for 2019. The difference in $\mathrm{POD_1}$ compared to MM in response to the localization of $f_\mathrm{T}$ and GS ranges between $\Delta\mathrm{POD_1}=(1.4-3.6)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $\Delta\mathrm{POD_1}=(2.7-7.7)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. A low soil water potential caused a reduction in $\mathrm{POD_1}$ in 2018 in cases when using PPFD0.8. This indicates that drought conditions can reduce the ozone damage risk. TODO: (With MM GS, $\mathrm{POD_1}$ values were higher in 2018 than in 2019. However, taking the localized GS into account, this is reversed for the subarctic temperature response function. This can be explained by more favorable growing conditions in 2019.)

For coniferous trees (Fig.~\ref{fig:pody_rel}b), most simulations slightly exceed the CL ($0-10\,\unit{mmol\,\chem{O_3}\,m^{-2}}$). Our localization does not show much difference compared to MM indicating the the MM Boreal Norway spruce parameterization was already well-adjusted to the local subarctic climate. We find a higher $\mathrm{POD_1}$ in 2019 than in 2018. The maximum difference in $\mathrm{POD_1}$ determined between localized GS for subarctic and cold parameterizations with respect to MM ranges between $(0.5-1.0)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(2.0-3.3)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. The maximum difference determined from PPFD0.8 (superscript) and PPFD1.2 (subscript) $f_\mathrm{light}$ amounts to $^{+1.0}_{-0.8}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ for 2018 and $^{+1.3}_{-1.0}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ for 2019.

For perennial grassland (Fig.~\ref{fig:pody_rel}c), all simulations with localize GS stay below the CL. Perennial grassland shows the smallest $\mathrm{POD_1}$, but a similarly large response to the temperature parameterization localizations as deciduous trees. Again, we find a reversal for predicted ozone risk in 2018 and 2019 for the subarctic type. Perennial grassland shows the lowest sensitivity to the light threshold. TODO: (and SWP is only relevant without localized GS (see above).) The overall difference in $\mathrm{POD_1}$ determined from the localized GS for subarctic and cold types with respect to MM ranges between $(2.2-3.4)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(4.4-6.4)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. The chosen lower maximum temperature in the subarctic parameterization probably caused the lower $\mathrm{POD_1}$ in 2018 (warm year) compared to the cold parameterization. This was not seen in 2019. The maximum difference in $\mathrm{POD_1}$ determined from the PPFD0.8 (superscript) and PPFD1.2 (subscript) variation of $f_\mathrm{light}$ amount to $^{+0.3}_{-0.2}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ and $^{+0.3}_{-0.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$, respectively.

The maximum difference in $\mathrm{POD_1}$ between simulations with the MM parameterizations and the localized temperature and light response functions is of the same order of magnitude as the difference between the two probed years. The magnitude of all described effects on $\mathrm{POD_1}$ varies between PFTs as well as years, but is by construction always larger than estimated with the MM parameterization. This means that the ozone damage risk could be possibly underestimated even in normal years and establishing dedicated subarctic parameterizations are a must to properly assess ozone damage risk in these regions.

@@

We estimated the biomass reductions in accordance to the CLs \citep{ICP:MappingManual2017, ESPR:Hayes2021} for the most extreme temperature parameterization localization (\emph{subarctic}) for each PFT. The estimates for SWP=off are listed in Table~\ref{tab:biomass_reduction}. We defined an uncertainty range from the difference in biomass reduction caused by the variation of $f_\mathrm{light}$. Superscript refers to PPFD0.8 and subscript to PPFD1.2. The reported range corresponds to the effect of taking SWP into account or not. As SWP reduces $\mathrm{POD_1}$, biomass reduction is decreased when taking SWP into account. The biomass reduction ranges between $2.5\,\unit{\%}$ (coniferous, 2018) and $17.4\,\unit{\%}$ (deciduous, 2019). For deciduous trees this amounts to a reduction by up to $55\,\unit{\%}$ compared to MM. From an agro-economical point of view in our focus area, where the grass is cut only once a year, a loss of biomass of more than $10\,\unit{\%}$ can have severer consequences compared to more productive areas in central Europe. Therefore, CL may also need localization.
Coniferous trees are the least affected PFT and the magnitude of total biomass reduction is almost independent of the choice of the parameterization within given uncertainties.

\begin{table}[t]
  \caption{Estimated total biomass reduction in \unit{\%} for temperature acclimation \emph{subarctic} with localized GS and SWP=off. The uncertainty ranges reported here correspond to maximum a divergence deducted from varying $f_\mathrm{light}$ and SWP=on. For comparison, the corresponding biomass reduction for the MM parameterization averaged over both years is shown. In this case, standard deviation is computed from all sensitivity simulations ($\pm 20\,\unit{\%}$ in extent of stomatal opening at low light, SWP on/off, and localized GS).}
  \label{tab:biomass_reduction}
\begin{tabular}{lllll}
\tophline
Year & \multicolumn{4}{c}{PFT}\\
& Deciduous tree& Coniferous tree & \multicolumn{2}{c}{Perennial grassland}\\
& & & total & above ground\\
\middlehline
2018 & $15.5^{+(1.9...2.1)}_{-(0.8...1.5)}$ & $2.5^{+0.2}_{-0.2}$ & $9.7^{+0.2}_{-0.2}$ & $12.4^{+0.2}_{-0.2}$\\
2019 & $17.4^{+2.5}_{-1.8}$ & $3.0^{+0.2}_{-0.3}$ & $10.5^{+(0.1...0.2)}_{-(0.0...0.1)}$ & $13.7^{+(0.1...0.3)}_{-(0.1...0.3)}$\\
\middlehline
$\left<\mathrm{MM}\right>$ & $11.2\pm 1.1$ & $2.31\pm 0.04$ & $7.5\pm 0.9$ & $9.6\pm 1.4$\\
\bottomhline
\end{tabular}
%\belowtable{$^*$ above ground biomass} % Table Footnotes
\end{table}
